---
title: 'DATA 605: Week 13 Assignment'
author: "Aaron Grzasko"
date: "4/26/2017"
output: 
    html_document:
        theme: default
        highlight: haddock
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

## Numerical Integration and Differentiation  
<br>  

#### Derivative Function  

**Write a program to compute the derivative of $f(x) = x^3 + 2x^2$ at any value of x. Your function should take in a value of x and return back an approximation to the derivative of f(x) evaluated at that value. You should not use the analytical form of the derivative to compute it. Instead, you should compute this approximation using limits.**  

Our function will return an approximation based on the following limit definition of the derivative:  

$$f'(x)=\lim_{\Delta x\to0} \frac{f(x + \Delta x)-f(x)}{\Delta x}$$  
The key idea with this function is to use a very small value for $\Delta x$, as the approximation improves as the size of $\Delta x$ decreases.

```{r}
# general function to approximate derivative
# inputs:  f(x), x, delta x
d.x <- function(func, x, delta = 10^-6){
  (func(x + delta) - func(x)) / delta
}
```

Let's calculate the derivative of $f(x) = x^3 + 2x^2$ at multiple values of $x$ using our homegrown derivative function:  
```{r}
# function of interest
myfunc1 <- function(x) x^3 + 2*x^2

# calculate derivatives for x = 1 to 5
my_deriv <- sapply(X=1:5, d.x, func=myfunc1)
my_deriv
```

Now we'll compare our homegrown derivative calculations against the values produced by R's `D()` function, which is included in the *stats* package.  

```{r}
# derivative of f(x) = x^3 + 2x^2 based on D() function
d.x_R <- function(x){} 
body(d.x_R) <- D(body(myfunc1),'x')

# derivative output for x values 1 to 5
R_deriv <- sapply(1:5, d.x_R)
R_deriv

# verify homegrown function output is approx. equal to D()'s output
abs(my_deriv - R_deriv) < 0.0001

```
  
#### Integral Function  
  
**Now, write a program to compute the area under the curve for the function $3x^2+4x$ in the range $x = [1, 3]$. You should first split the range into many small intervals using some really small $\Delta x$ value (say 1e-6) and then compute the approximation to the area under the curve.**  
  
Our function will be based on the following definition of the integral:  

$$\int f(x)dx= \lim_{\Delta x \to 0} \sum f(x) \Delta x$$  
Once again, the key to a reasonable approximation is to use a small value for $\Delta x$ (e.g. 1e-6).  
  
Below, we produce two functions to approximate the integral:  a loop-based version and a vectorized version.   

```{r}
# general function to approximate definite integral
# inputs: f(x), lower limit, upper limit, delta x

s.x_loopy <- function(func, l, u, delta = 10^-6){
  output <- 0
  for (x in seq(l,u, by=delta)){
    output <- func(x)*(delta) + output
  }
  output
}

# vectorized version of function to approximate integral
# same inputs as above
s.x_vec <- function(func, l,u, delta = 10^-6){
  sum(func(seq(l,u, by=delta))*delta)
}

```

Let's calculate the integral of $3x^2 +4x$ for both homegrown functions using lower limit x = 1, and upper limit x=3:  
```{r}
# function of interest
myfunc2 <-function(x) 3*x^2 + 4*x

# loopy integral output 
my_int_l <- s.x_loopy(myfunc2,1,3)
my_int_l

# vectorized integral output
my_int_v <-s.x_vec(myfunc2,1,3)
my_int_v
```

Finally, let's check our work using R's `integrate()`, another function which is featured in the *stats* package.  
 
```{r}
# integral using R built-in function
s.x_R <- integrate(myfunc2, 1,3)
s.x_R

# verify s.x_loopy produces same output as s.x_R  
abs(my_int_l - s.x_R$value) < 0.0001

# verify s.x_vec produces same output as s.x_R  
abs(my_int_v - s.x_R$value) < 0.0001

```

## Analytical Integration and Differentiation  
  
**Please solve these problems analytically (i.e. by working out the math) and submit your answers.**  

**1. Use integration by parts to solve for** $\mathbf{\int sinx~cosx~dx}$


Integration by parts is not the preferred method for solving this integral: splitting the original expression into component parts includes the original term to be integrated as well as a new term.  The two possibilities are:   

$$\int sinx~cosx~dx=\int sin^2x - \int sinx~cosx~dx$$  
  
-or-  

$$\int sinx~cosx~dx= - cos^2x- \int sinx~cosx~dx$$

Instead, we'll solve using *u* substitution.  Let $u = sinx$, $du = cosx~dx$:  
 
Plug terms above into the original expression:  
  
$$\int u~du$$
$$=\frac{u^2}{2}+C$$ 
Replace $u$ with $sinx$:  
$$=\frac{sin^2x}{2}+C$$
  

**2. Use integration by parts to solve for $\mathbf{\int x^2~e^x~dx}$**  
  
Let *$u=x^2$ , $du=2x$, $dv=e^x$, $v = e^x$*  
  
$$\int x^2~e^x~dx=x^2e^x-2\int x~e^xdx$$  

Now, use integration by parts for the second expression on the right hand side of the equation above.

Let *$u=x$ , $du=$, $dv=e^x$, $v = e^x$*  
$$\Rightarrow x^2e^x-2(xe^x-\int e^xdx) $$  
$$=x^2e^x-2xe^x+2~e^x + C$$  
$$=e^x(x^2-2x+2)+C$$  

**3.  What is $\mathbf{\frac{d}{dx}(x~cosx)}$?**  

Use the product rule for derivatives:  
  
$$\frac{d}{dx}f(x)g(x)=f'(x)g(x)+f(x)g'(x)$$  

$$\frac{d}{dx}(x~cosx) = cosx - x~sinx$$  
  
**4. What is $\mathbf{\frac{d}{dx}{e^{x^4}}}$?**  

Use the chain rule:  
  
$$\frac{df}{dx}=\frac{df}{du}~\frac{du}{dx}$$  

Let $u=x^4$. 

$$\frac{df}{du}=e^u$$  
$$\frac{du}{dx}=4x^3$$

$$\Rightarrow \frac{d}{dx}e^{x^4}=4x^3~e^{x^4}$$  


### References 

* How to use `D()` function to calculate derivatives:  <http://rpubs.com/wkmor1/simple-derivatives-in-r>
